An Accurate Model for SARS-CoV-2 pooled RT-PCR Test Errors

### 2021-03-30 comment.tex
figure 1 caption, a lot of changes in the discussion

\begin{figure}[H]
  \centering
  \begin{tikzpicture}
    \begin{scope}[every node/.style={circle,thick,draw}]
      \node[shape=circle,draw=black] (A) at (0,0) {$\theta_1=1$};
      \node[shape=circle,draw=black] (B) at (2,0) {$\theta_2=0$};
      \node[shape=circle,draw=black] (D) at (4,0) {$\theta_3=0$};
      \node[shape=circle,draw=black] (E) at (6,0) {$\theta_4=1$};

      \node[shape=ellipse,draw=black]
      (F) at (9,0) {False-positive};

      \node[shape=rectangle,draw=black] (G) at (4,7)
           {$\Pr(\data=0|\theta)=(1-\sens)^2\spec$};
    \end{scope}

    \begin{scope}[>={Stealth[black]},
        every node/.style={fill=white},
        every edge/.style={draw=red,very thick}]
      \path [->] (A) edge node {$1-\sens$} (4.5,6.6);
      %\path [->] (B) edge node {$0$} (G);
      %\path [->] (D) edge node {$0$} (G);
      \path [->] (E) edge node {$1-\sens$} (4.5,6.6);
      \path [->] (F) edge node {$\spec$} (5.5,6.6); 
    \end{scope}
  \end{tikzpicture}
  \caption{Illustration of our model. A pool contains individuals
    $\{1,2,3,4\}$ with state $\theta=(1,0,0,1)$ (i.e. individuals 1
    and 4 are infected). A negative pooled test ($\data=0$) implies
    that three detection paths failed. A false-negative occurred for
    $1$ and $4$, each with probability $(1-\sens)$ for
    each. Additionally, no erroneous detection of SARS-CoV-2 occurred,
    with probability $\spec$. Individuals $2$ and $3$ are not infected
    and do not contribute to the probability of the pooled test
    result. A positive pooled test arises if any one of the above
    mentioned paths results in a detection.}\label{fig:likelihood}
\end{figure}

@@

\section*{Discussion}
%\subsection{Short recap}
We have presented a novel model for SARS-CoV-2 pooled tests. To the best of our knowledge, ours is the
only one that is actually based on experimental data.

we have shown that low values of infection prevalence, or
low values of single test false-negative rates, increase
$\err{exact}$. These results remain qualitatively similar under varying
parameter values, in the observed ranges
\cite{KitComparison,EstimatingRatesKucrika, EstimatingRatesLourenco,
  InterpretingCOVID19Test} (Figure \ref{fig1}).

Our results lead to a fact almost disregarded in
\cite{DorfmanYuvalDor}: although (Dorfman) pooling is most efficient
when prevalence is low, such circumstances are exactly those leading
to a substantial increase in false-negative rates. Such behavior of
pooling is not implied by the common assumption

%\subsection{Implications for group testing in different prevalence settings}
Our results have practical implications for the use of pooling
schemes. As the COVID-19 pandemic progresses, the infection prevalence
in various tested populations undergoes frequent changes. Hence, as
our results suggest, pooling strategies for mass testing should be
used with caution in populations where infection rates are low. Such
mass-tested populations often include air travel passengers
\cite{JTM,RobinHood}, potentially importing infections, or
asymptomatic individuals, such as children, and can be crucial for
controlling outbreaks \cite{MinaScience}.

@@

%\subsection{Ways to reduce errors}
Although our application focused on Dorfman pooling, our analysis
applies to other pooling schemes. Pooling schemes
(e.g. \cite{MatrixPooling,Lion, Kim}), require some sequence of
positive pooled results to correctly identify Donald as
infected. Consider the hierarchical pooling scheme \cite{Lion, Kim}:
If the first pool yields a positive result, it is split in two. Then
the splitting is repeated until resulting pools are negative or
individuals are tested separately. With an initial pool size of 32,
Donald will necessarily have to test positive in pools of size 32, 16,
8, 4 and 2, as well as in a single test, for the scheme to correctly
identify him as infected. Compare this to the Dorfman scheme that
requires a positive test in a pool of size $N=8$, and an additional
single positive test to identify Donald as infected. The hierarchical
pooling scheme of \cite{Lion, Kim} will necessarily yield more
false-negatives than Dorfman pooling --- there are additional places
for it to fail.

Thus, finding better pooling schemes is imperative. One scheme that
holds great promise is DOPE \cite{DOPE}. This scheme utilizes
considerably less RT-PCR tests and has lower error rates than many
competing schemes.

@@

As mentioned in \cite{DorfmanYuvalDor}, introducing a positive
dependence within a pool decreases the false-positive rate. In the
extreme case, consider a fully connected pool, where one infection
implies the entire pool is infected. In this case, a calculation
analogous to the one conducted above recovers the initial
false-negative rate $1-\Se$. Interestingly, pooling was also noted to
have increased throughput when infection probabilities are dependent
between the pooled individuals \cite{DorfmanYuvalDor}, providing
another advantage to sampling dependent individuals in pooling
schemes.

To conclude, pooling is an important technique which can improve
testing throughput in a cost-effective manner. Nevertheless, care must
be given to pooling schemes' false-negative rates and
efficiency. These are analyzed via a common assumption that we refute.
The implications of refuting the common assumption were presented and
should be considered by practitioners.

>>>>

\begin{figure}[H]
  \centering
  \begin{tikzpicture}
    \begin{scope}[every node/.style={circle,thick,draw}]
      \node[shape=circle,draw=black] (A) at (0,0) {$\theta_1=1$};
      \node[shape=circle,draw=black] (B) at (3,0) {$\theta_2=1$};
      \node[shape=circle,draw=black] (D) at (6,0) {$\theta_3=0$};
      %\node[shape=circle,draw=black] (E) at (6,0) {$\theta_4=1$};

      \node[shape=ellipse,draw=black] (F) at (8,5) {False-positive};

      \node[shape=rectangle,draw=black] (G) at (1.5,5)
           {$\Pr(\data=0|\theta)=(1-\sens)^2\spec$};
    \end{scope}

    \begin{scope}[>={Stealth[black]},
        every node/.style={fill=white},
        every edge/.style={draw=black,very thick}]
      \path [->] (A) edge node {$1-\sens$} (2,4.6);
      %\path [->] (B) edge node {$0$} (G);
      %\path [->] (D) edge node {$0$} (G);
      \path [->] (B) edge node {$1-\sens$} (2.5,4.6);
        \path [->] (F) edge node {$\spec$} (3.7,5); 
    \end{scope}
  \end{tikzpicture}
  \caption{Illustration of our model. A pool contains individuals
    $\{1,2,3\}$ with state $\theta=(1,1,0)$ (i.e., individuals 1
    and 2 are infected). A negative pooled test ($\data=0$) occurs when
    three detection paths fail: A false-negative occurs for individuals
    $1$ and $2$, each with probability $(1-\sens)$.
    Additionally, no false-positive detection of SARS-CoV-2 occurs,
    with probability $\spec$. Individual $3$ is not infected
    and does not contribute to the probability of the pooled test
    result. The pooled test is positive if any path results in a detection.}\label{fig:likelihood}
\end{figure}

@@

\section*{Discussion}
%\subsection{Short recap}
Here, we presented a novel probabilistic model for the outcomes of pooled PCR tests, parametrized for SARS-CoV-2. Contrary to previous models, we assume, based on data (see section XX), that multiple infected individuals increase the chance of a positive pool.
A direct consequence of our model is that  false-negative rates depend on infection prevalence. Specifically, low values of infection prevalence increase
$\err{exact}$. These results remain qualitatively similar under varying
parameter values, in the observed ranges
\cite{KitComparison,EstimatingRatesKucrika, EstimatingRatesLourenco,
  InterpretingCOVID19Test} (Figure \ref{fig1}).

Our results lead to a conflict: low infection prevalence leads to high efficiency of Dorfman pooling \cite{DorfmanYuvalDor}, while also increasing false-negative rates.

fact almost disregarded in
\cite{DorfmanYuvalDor}: although (Dorfman) pooling is most efficient
when infection prevalence is low, such circumstances are exactly those leading
to a substantial increase in false-negative rates. Such behavior of
pooling is not implied by the common assumption

%\subsection{Implications for group testing in different prevalence settings}
As the COVID-19 pandemic progresses, the infection prevalence
in various tested populations undergoes frequent changes. Hence, as
our results suggest, pooling strategies employed for mass testing should be
used with caution in populations where infection rates are low. Such
mass-tested populations often include air travel passengers
\cite{JTM,RobinHood}, potentially importing infections, or
presymptomatic and asymptomatic individuals, and can be crucial for
controlling outbreaks \cite{MinaScience}.

@@

%\subsection{Ways to reduce errors}
Although our application focused on Dorfman pooling, our analysis
applies to other pooling schemes. Pooling schemes
(e.g. \cite{MatrixPooling,Lion, Kim}), require some sequence of
positive pooled results to correctly identify Donald as
infected. Consider the hierarchical pooling scheme \cite{Lion, Kim}:
If the first pool yields a positive result, it is split in two; otherwise, everyone are declared negative. Then
the splitting is repeated until resulting pools are negative or
individuals are tested separately. With an initial pool size of 32,
Donald will necessarily have to test positive in pools of size 32, 16,
8, 4 and 2, as well as in a single test, for the scheme to correctly
identify him as infected. Compare this to the Dorfman scheme that
requires a positive test in a pool of size $N=8$, and an additional
single positive test to identify Donald as infected. The hierarchical
pooling scheme of \cite{Lion, Kim} will necessarily yield more
false-negatives than Dorfman pooling --- there are additional places
for it to fail.

Thus, finding improved pooling schemes is imperative. An especially important consideration in designing such schemes is explicitly taking the intrinsic PCR error rates into account. We (ref), as well as others (ref), have developed such models, although their implementation in real-world settings has not yet commenced.

@@

Another important factor in pooling is dependence between samples. As mentioned in \cite{DorfmanYuvalDor}, introducing a positive
dependence within a pool decreases the false-positive rate. In the
extreme case, consider a fully connected pool, where one infection
determines the entire pool is infected. In this case, a calculation
analogous to the one conducted above (section??) recovers the initial
false-negative rate $1-\Se$. Interestingly, pooling was also noted to
have increased throughput when infection probabilities are dependent
between the pooled individuals \cite{DorfmanYuvalDor}, providing
another advantage to sampling dependent individuals in pooling
schemes. Hence, deliberately pooling individuals with positive dependence of infection should be further studied and likely encouraged when performing pooling.

To conclude, pooling is an important technique that can improve
testing throughput in a cost-effective manner. Nevertheless, care must
be given to pooling schemes' false-negative rates, especially under low infection prevalence settings.
